Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-02 papers/halfdisk/halfdisk.tex
Edits to paper

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mleftright}
\usepackage{todonotes}

@@

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align*}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align*}

@@

\begin{proof}
First, note that, using a change of variable,
\begin{align*}
	\norm{\hdopnkab}_\Wab^2 &= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(y)^2 \: w_P^{(b)}(y) \: \D y \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align*}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align*}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align*}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align*}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align*}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.

$J_x^{(a,b)}, J_y^{(a,b)}$ have banded-block-banded structures. A block-banded matrix is a block $n$-diagonal matrix where $n$ is reasonably small for the size of the full matrix. A banded-block-banded matrix is one where the blocks themselves are banded matrices. $J_x^{(a,b)}, J_y^{(a,b)}$ are given by:
\begin{align*}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align*}
where
\begin{align*}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align*}

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align*}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bighdopab(x,y)
=
\begin{pmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align*}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv \genjac_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:
\begin{align*}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+2)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align*}

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{n+2}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align*}
\begin{cases}
\bighdopab_{-1}(x,y) := 0 \\
\bighdopab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bighdopab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bighdopab_n(x,y) - \Dnt C_n  \,\bighdopab_{n-1}(x,y), \quad n = 1,2,3,\dots.
\end{cases}
\end{align*}

@@

\begin{definition}\label{def:sparsityofdifferentialoperators}
Define the operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ according to:
\begin{align*}
{\partial f \over \partial x}&= \bighdop^{(a+1,b+1)}(x,y)^\top \: D_x^{(a,b)} \: \mathbf{f}, \\
{\partial f \over \partial y} &= \bighdop^{(a,b+1)}(x,y)^\top \: D_y^{(a,b)} \: \mathbf{f}, \\
{\partial \over \partial x}[\Wab(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1)}(x,y)^\top \: W_x^{(a,b)} \: \mathbf{f}, \\
{\partial \over \partial y}[\Wab(x,y) \: f(x,y)] &= \bigW^{(a,b-1)}(x,y)^\top \: W_y^{(a,b)} \: \mathbf{f}.
\end{align*}
\end{definition}

@@

\begin{lemma}\label{lemma:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ from Definition \ref{def:sparsityofdifferentialoperators} are sparse, with banded-block-banded structure.
\end{lemma}

\begin{proof}
We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$, we can also expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^2_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $\tilde{y} = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} \\
	&= \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
Now, integrating by parts, noting that the weight $w_P^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} &= \ip< P_k^{(b,b) \: \prime}, \: w_P^{(b+1)} \: P_{j}^{(b+1,b+1)} >_1 \\
	&= \ip< P_k^{(b,b)}, \: \ddy [w_P^{(b+1)} \: P_{j}^{(b+1,b+1)}] >_1 \\
	&= \ip< P_k^{(b,b)}, \: P_{j}^{(b+1,b+1) \: \prime} >_{w_P^{(b+1)}} - 2 b \ip< P_k^{(b,b)}, \: y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} \\
	&= \ip< P_k^{(b,b)}, \: w_P^{(1)} \: P_{j}^{(b+1,b+1) \: \prime} -2b \:y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} 
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,k-1} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}},
\end{align*}
which is zero for $m < n-1$. Finally, we simply have that
\begin{align*}
	c_{n-1,k-1} &= \frac{1}{\normjac^{(b)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage[margin=10pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mleftright}
\usepackage{todonotes}

@@

The weight $w_P(x)$ is an ultraspherical weight, and the corresponding OPs are the Jacobi polynomials  $\{P_n^{(b, b)}\}$. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. Thus we arrive at a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align*}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align*}

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, first note that, using a change of variable $t = \frac{y}{\rho(x)}$,
\begin{align*}
	\norm{\hdopnkab}_\Wab^2 &=  \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \Wab(x,y) \: \D y \: \D x \nonumber \\
	&= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b+k)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(t)^2 \: w_P^{(b)}(t) \: \D t \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align*}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\) using the same change of variable:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= 
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}.
\end{align*}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align*}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align*}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a function's expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \mapsto x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.

$J_x^{(a,b)}, J_y^{(a,b)}$ have banded-block-banded structures. A block-banded matrix is a block $n$-diagonal matrix where $n$ is reasonably small for the size of the full matrix. A banded-block-banded matrix is one where the blocks themselves are banded matrices. $J_x^{(a,b)}, J_y^{(a,b)}$ are given by:
\begin{align*}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align*}
where
\begin{align*}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber
\end{align*}
\begin{align*}
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align*}

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align*}
\renewcommand\arraystretch{1.3}
\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{pmatrix}
\bighdopab(x,y)
=
\begin{pmatrix}
	\hdopab_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align*}
where we note $\hdopab_{0,0}(x,y) \equiv \genjac_0^{(a,b)} \: P_0^{(b,b)}$, and for each $n = 0,1,2\dots$:
\begin{align*}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+2)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align*}

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{n+2}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence:
\begin{align*}
\begin{cases}
\bighdopab_{-1}(x,y) := 0 \\
\bighdopab_{0}(x,y) := P^{(a,b)}_{0,0} \\
\bighdopab_{n+1}(x,y) = -\Dnt (B_n-G_n(x,y)) \bighdopab_n(x,y) - \Dnt C_n  \,\bighdopab_{n-1}(x,y), \quad n = 0,1,2,\dots.
\end{cases}
\end{align*}

@@

\begin{definition}\label{def:differentialoperators}
Define the operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ according to:
\begin{align*}
{\partial f \over \partial x}&= \bighdop^{(a+1,b+1)}(x,y)^\top \: D_x^{(a,b)} \: \mathbf{f}, \\
{\partial f \over \partial y} &= \bighdop^{(a,b+1)}(x,y)^\top \: D_y^{(a,b)} \: \mathbf{f}, \\
{\partial \over \partial x}[\Wab(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1)}(x,y)^\top \: W_x^{(a,b)} \: \mathbf{f}, \\
{\partial \over \partial y}[\Wab(x,y) \: f(x,y)] &= \bigW^{(a,b-1)}(x,y)^\top \: W_y^{(a,b)} \: \mathbf{f}.
\end{align*}
\end{definition}

@@

\begin{lemma}\label{lemma:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure.
\end{lemma}

\begin{proof}
We proceed with the case for the operator $D_y^{(a,b)}$ for partial differentiation by $y$. Since $\{\hdop^{(a, b+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: \hdop^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
\begin{align*}
	c_{m,j} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{m,j}}^2_{W^{(a, b+1)}}}.
\end{align*}
Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,j} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \rho(x)^{k+j} \: \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} \\
	&= \ip< \genjacnmk^{(a,b+k+\half)}, \genjacmmj^{(a,b+1+j+\half)} >_{\genjacw^{(a,b+1+\half(k+j))}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
Now, integrating by parts, noting that the weight $w_P^{(b)}$ is a polynomial degree $2b$ and vanishes at the limits of the integral for positive parameter $b$, we have that
\begin{align*}
	\ip< P_k^{(b,b) \: \prime}, \: P_{j}^{(b+1,b+1)} >_{w_P^{(b+1)}} &= \ip< P_k^{(b,b) \: \prime}, \: w_P^{(b+1)} \: P_{j}^{(b+1,b+1)} >_1 \\
	&= \ip< P_k^{(b,b)}, \: \ddy [w_P^{(b+1)} \: P_{j}^{(b+1,b+1)}] >_1 \\
	&= \ip< P_k^{(b,b)}, \: P_{j}^{(b+1,b+1) \: \prime} >_{w_P^{(b+1)}} - 2 b \ip< P_k^{(b,b)}, \: y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} \\
	&= \ip< P_k^{(b,b)}, \: w_P^{(1)} \: P_{j}^{(b+1,b+1) \: \prime} -2b \:y \: P_{j}^{(b+1,b+1)} >_{w_P^{(b)}} 
\end{align*}
which is zero for $j < k-1$. Further,
\begin{align*}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{m,k-1} >_{W^{(a, b+1)}} &= \ip< \genjacnmk^{(a,b+k+\half)}, \genjac_{m+1-k}^{(a,b+k+\half)} >_{\genjacw^{(a,b+k+\half)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}},
\end{align*}
which is zero for $m < n-1$. Finally, we simply have that
\begin{align*}
	c_{n-1,k-1} &= \frac{1}{\normjac^{(b)}} \: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}.
\end{align*}
